Naturalness and Long-lived Beyond the Standard Model Particles

### 2017-07-29 tex/draft.tex
Edit through displaced vertices (inclusive)

\section{LHC Searches for Long-lived Particles}
    \fxnote{Include diagrams for signal models} 
    \fxnote{mention which data was used for each search} 
    \fxnote{Steal limit plots and any cool figures from papers} 
\subsection{Overview}
    LLPs leave distinct signatures that differentiate signal from SM background but also add complexity to LLP analyses. CMS and ATLAS were designed to reconstruct stable particles from prompt decays, so LLP searches are often forced to stray from the traditional analysis path in terms of object reconstruction\fxnote{really reconstruction? Maybe object definition?} and background estimation. This report focuses on four complementary LLP signatures: displaced vertices, disappearing tracks, slow-moving particles, and particles that stop completely before decaying.

    Displaced vertices and disappearing tracks both look for\fxnote{reword. the vertices and tracks don't look for anything} particles that are unstable on detector timescales. Displaced vertex searches are sensitive to LLPs that decay to visible particles in the inner volume \fxnote{specify} of the detector and leave tracks that can be traced back to a vertex that is sufficiently displaced from the nominal interaction point \fxnote{wordy}. Disappearing track searches are sensitive to LLPs that decay to particles that are either neutral or too soft to be reconstructed. Together, these two types of analyses cover \fxnote{something something} parameter space of LLPs that decay on detector timescales.

    Slow and stopped particle searches are both sensitive to LLPs that are detector-stable but far more massive than any stable SM particle. Searches for low-velocity particles use ionization energy loss per unit distance (dE/dX) and time-of-flight (TOF) measurements to look for particles that are moving more slowly than the expected SM particles \fxnote{redundant}, which normally \fxnote{quantify} have $\beta=\frac{v}{c} > \num{0.9}$. Stopped particle searches take this idea one step further and look for particles that actually come to rest inside the detector before decaying, a phenomenon that could be possible with long lifetimes and small initial kinetic energy. These analyses look for calorimeter energy deposits or hits in the muon system that are out of time with proton bunch crossings that could be due to the decay of stopped particles.

@@

\subsection{Displaced Vertices}
    \fxnote{cite CONF-2017-026}
    LLPs that decay before reaching the inner tracker but still live longer than the vast majority of SM particles can be identified with a displaced vertex (DV). That is, a reconstructed vertex far from the nominal interaction point. ATLAS performs a search for such particles by looking for at least one DV along with large MET and large track multiplicity. The search is sensitive to LLPs with lifetimes in the \SI{20}{\pico\s} to \SI{10}{\nano\s} range.

    This search uses a simplified model in the split SUSY paradigm as a benchmark. In the simplified model, gluinos are kinematically accessible while squarks are not. The relevant process, then, is pair produced gluinos that decay to a quark and virtual squark, which then decays to a quark and the LSP, which is presumed to be a neutralino \fxnote{steal diagram from andrew's preapproval. I can't access it - ask andrew}. Because the squark mass is so high, the gluino lives long enough to form an R-hadron \fxnote{first use. Define here if not in LLP section} with SM partons and propogate some measureable distance away from the IP before decaying.

    Jets and MET are reconstructed as in standard analyses, but special care must be taken when looking for displaced tracks. The standard ATLAS tracking algorithm constrains the transverse and longitudinal impact parameters of potential tracks, so the authors employ an additional algorithm, referred to as large-radius tracking (LRT). LRT is run after the standard tracking algorithm and uses only hits that were not used in standard track reconstruction. By relaxing d0, z0,\fxnote{define} and number of hits requirements, LRT is able to reconstruct tracks that point back to DVs.

    Because no SM process will produce events that meet the mass and track multiplicity requirements on the DV, only instrumental backgrounds are considered. Specifically, the authors consider hadronic interactions in detector material, incorrectly combined vertices from short-lived SM particles, and incorrectly added hits from unrelated tracks to low-mass vertices.

    The authors minimize the effect of hadronic interactions by vetoing the material-dense regions of the detector and estimate the residual background by extrapolating from the low-mass region, which is dominated by hadronic interactions. After a dedicated study, the background from merged vertices was found to be negligable. Finally, the background from crossing tracks is determined with a model that adds pseudo-tracks to data vertices and is normalized by comparing the sample of 3-track vertices in data to (2+1)-track vertices in the model.

>>>>

\section{LHC Searches for Long-lived Particles}
    \fxnote{Include diagrams for signal models} 
    \fxnote{mention which data was used for each search} 
    \fxnote{Steal limit plots and any cool figures from papers} 
\subsection{Overview}
    LLPs leave distinct signatures that differentiate them from SM backgrounds but also add complexity to LLP analyses. CMS and ATLAS were mainly designed to reconstruct stable particles from prompt decays, so LLP searches are often forced to stray from the traditional analysis path in terms of object reconstruction\fxnote{really reconstruction? Maybe object definition?} and background estimation. This report focuses on four complementary LLP signatures: displaced vertices, disappearing tracks, low-velocity particles, and particles that stop completely before decaying.

    Displaced vertex and disappearing track analyses both look for particles that are unstable on detector timescales. Displaced vertex searches are sensitive to LLPs that decay to visible particles within the inner volume \fxnote{specify} of the detector, leaving tracks that point back to a vertex some measurable distance from the nominal interaction point (IP). Disappearing track searches are sensitive to LLPs that decay to particles that are either neutral or too soft to be reconstructed. Together, these two types of analyses cover a wide range of lifetimes and topologies for LLPs that decay on detector timescales\fxnote{dangerous as hell. Be really good at giving a quantitative explanation of this sentence}.

    Slow and stopped particle searches, on the other hand, are sensitive to detector-stable\fxnote{define?} LLPs that are far more massive\fxnote{be able to quantify} than any stable SM particle. The lower velocity of these high-mass particles shows up as increased ionization energy loss per unit distance ($\langle\frac{dE}{dx}\rangle$) and longer time-of-flight (TOF) measurements than SM particles, which generally have $\beta > \num{0.9}$. Stopped particle searches take this idea one step further and look for particles that actually come to rest inside the detector before decaying, a phenomenon that could be possible with long lifetimes and small initial kinetic energy. These analyses look for calorimeter energy deposits or hits in the muon system that are out of time with proton bunch crossings that could be due to the decay of stopped particles.

@@

\subsection{Displaced Vertices}
    \fxnote{cite CONF-2017-026}
    LLPs that live longer than most SM particles but decay before reaching the inner tracker can be identified with a displaced vertex (DV)\fxnote{DVs defined clearly enough in overview?}. ATLAS performs a search for such particles by looking for at least one DV along with large MET and large track multiplicity.\fxnote{make sure I can explain why met and multiplicity} The search is sensitive to LLPs with lifetimes in the \SI{20}{\pico\s} to \SI{10}{\nano\s} range\fxnote{Am I defining sensitivity range consistently across analyses?}.

    This search\fxnote{repetative wording} uses a simplified benchmark model in the split SUSY paradigm where gluinos are kinematically accessible but squarks are not. The relevant process, shown in \fxnote{reference figure}, is pair-produced gluinos that decay to a quark-\fxnote{how to punctuate?}virtual squark pair. The virtual squark then decays to a quark and the LSP, presumed to be a neutralino\fxnote{have I defined neutralino?}\fxnote{why is this a reasonable assumption?} \fxnote{steal diagram from andrew's preapproval. I can't access it - ask andrew}. The large mass of the intermediate squark suppresses the gluino decay, giving the gluino enough time to form an R-hadron\fxnote{first use. Define here if not in LLP section} that propagates some measureable distance before decaying.

    Jets and MET are reconstructed as in conventional analyses, but dispaced vertices demand special care when identifying tracks. The standard ATLAS tracking algorithm constrains the transverse and longitudinal impact parameters ($\mathrm{d}_0$ and $\mathrm{z}_0$, respectively) of potential tracks, so the authors employ an additional algorithm, referred to as large-radius tracking (LRT). LRT is run after the standard tracking algorithm and uses only hits that were not used in standard track reconstruction. By relaxing requirements on $\mathrm{d}_0$, $\mathrm{z}_0$, and number of hits, LRT is able to reconstruct tracks that point back to DVs.

    Because no SM process produces events that meet the mass and track multiplicity requirements on the DV, only instrumental backgrounds are considered. Specifically, the authors consider hadronic interactions in detector material, incorrectly combined vertices from short-lived SM particles, and hits from unrelated tracks that are incorrectly added to low-mass vertices\fxnote{unnecessary details? I need to get better at explaining this} .

    The authors minimize the effect of hadronic interactions by vetoing the material-dense regions of the detector\fxnote{cool figure to steal} and estimate the residual background by extrapolating from the low-mass region, which is dominated by hadronic interactions. After a dedicated study, the background from merged vertices was found to be negligable. Finally, the background from crossing tracks is determined with a model that adds pseudo-tracks to data vertices and is normalized by comparing the sample of 3-track vertices in data to (2+1)-track vertices in the model.
